---
title: "Introduction"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| eval: true

pmetrics <- function(){
    knitr::asis_output("[Pmetrics]{style=\"color: #841010; font-family: 'Arial', Arial, sans-serif; font-weight: 900;\"}")
}
```

Thank you for your interest in `r pmetrics()`! This online book provides
instructions and examples to assist users of the `r pmetrics()` R package, by
the Laboratory of Applied Pharmacokinetics and Bioinformatics at the
Saban Research Institute, Children's Hospital Los Angeles, and the
Department of Pediatrics, Keck School of Medicine, University of
Southern California. Please see our website at
[[http://www.lapk.org]{.underline}](http://www.lapk.org) for more
information.

## Disclaimer

You, the user, assume all responsibility for acting on the results
obtained from `r pmetrics()`. The Laboratory of Applied Pharmacokinetics and
Bioinformatics (LAPKB), members and consultants to LAPKB, and Children's
Hospital Los Angeles and the University of Southern California and their
employees assume no liability whatsoever. Your use of the package
constitutes your agreement to this provision.

## Installing Pmetrics

`r pmetrics()` and all required components will run under MacOS (Unix), Windows,
and Linux. Instructions on obtaining and installing the package are in the README section of our source code [repository in GitHub](https://github.com/LAPKB/Pmetrics_rust/).

## What This Guide Is Not

We assume that the user has familiarity with population modeling and R,
and thus this manual is not a tutorial for basic concepts and techniques
in either domain. We have tried to make the R code simple, regular and
well documented. A very good free online resource for learning the
basics of R can be found at [Stat Methods](http://www.statmethods.net/index.html). However, there are a number of other free or low-cost online resources which you can find with a simple or AI-assisted web search.

We recognize that initial use of a new software package can be complex,
so please feel free to contact us at any time, preferably through the
`r pmetrics()` [Disussions](https://github.com/LAPKB/Pmetrics/discussions). You can report [issues](https://github.com/LAPKB/Pmetrics/issues) online as well.

This manual is also not intended to be a theoretical treatise on the
algorithms `r pmetrics()` uses. For that, the user is directed to our
[website](http://www.lapk.org/publications.php).

## Getting Help and Updates

Within R, you can  use
`help("command")` or more simply just `?command` in the R console to see detailed help files for any `r pmetrics()` command. Many commands have examples included in this documentation and you can execute the examples with `example(command)`.

You can also type `PM_manual()` to launch this website from within Pmetrics. 

`r pmetrics()` will check for updates automatically every time you load it
with `library(Pmetrics)` and you are connected to the internet. 
If an update is available, it will provide a
brief message to inform you. You can then reinstall the package.

## Examples and Exercises

Included in this book are numerous code examples and exercises that will help you learn how to use `r pmetrics()` for R. They are designed to be read in sequence.

* Items that are [hyperlinked]() can be selected to cross reference
within this manual or link to external sites.

* `Items` correspond to inline examples of R code, which are not evaluated
in this document, but serve as templates for what may be typed into your 
R console or script. They may not necessarily be executable if typed verbatim.

* We have designed these pages with the idea that you can copy and paste the code chunks into your own scripts or Quarto documents to see how they work. You can hover over the code chunks to see a copy icon. Click that to copy the code to your clipboard so you can paste it into your own script or document. Alternatively, you can click the **Copy all code** button at the top right of every page to copy all the code chunks on that page to your clipboard.

**Note:** In many cases, we do not evaluate the code chunks because the ouput is too long, so you will need to run them in your own script or document.

```{r}
# Hover over the code chunk to see the copy icon.
```



