---
format:
  html:
     css: Extras/custom.scss
     include-in-header: Extras/copy-all-code.html
---



```{r}
#| label: setup
#| echo: false
#| message: false
#| eval: true
library(glue)
library(knitr)
library(Pmetrics)

pmetrics <- function(){
    knitr::asis_output("[Pmetrics]{style=\"color: #841010; font-family: 'Arial', Arial, sans-serif; font-weight: 900;\"}")
}


r_help <- function(pkg, name, fn = NULL) {
    if(is.null(fn)) {fn <- name}
    glue::glue("[`{name}`](https://rdrr.io/pkg/{pkg}/man/{fn}.html)")
}

gh_help <- function(name) {
    glue::glue("[`{name}`](https://lapkb.github.io/Pmetrics/reference/{name}.html)")
}
```

# Detailed example of a typical workflow

:::{.callout-tip}
Make sure you have run the [tutorial setup code](index.qmd#tutorial-setup) in your R session before copying, pasting and running example code here.
:::


## Load the Pmetrics library

First, remember you must always load the `r pmetrics()` library to access its functions in your R script or Quarto document.

```{r}
#| eval: false
#| echo: true
#| label: example-npag-1
library(Pmetrics)
```




## Create a run

To review, recall that the typical workflow in `r pmetrics()` is as follows:

1.  Create a `r gh_help("PM_data")` object.
2.  Create a `r gh_help("PM_model")` object.
3.  Fit the model to the data with the `$fit()` method of the model object. This returns a `r gh_help("PM_result")` object, which is also saved to the hard drive and can be loaded later with the `r gh_help("PM_load")` function.
4.  Examine the results with `r gh_help("PM_result")` methods and functions and revise as needed to improve the model fit. 

### Data objects

Since `r pmetrics()` always needs data and a model to run, let's create a new data object.

```{r}
#| eval: true
#| label: example-npag-4

exData <- PM_data$new(data = "Data/src/ex.csv")
```

`r pmetrics()` standardizes the data, checks for errors, and provides an error and attempted fix report fo any found.

You can look at a .csv data file directly by opening it in a spreadsheet program like Excel, or a text editor.

`exData` is an `r r_help("R6", "R6", "R6Class")` object, which means that contains both data and methods to process that data.

```{r}
#| label: example-npag-5
exData # prints the object, usually to the terminal, but in this case to a rendered "lite" spreadsheet
exData$data # contains your original data
exData$standard_data # contains the standardized and validated data
exData$summary() # prints the summary of the data to the terminal
```

Most `r pmetrics()` objects are R6 objects, so you can use the `$` operator to access their data and methods. Many of them have a `summary()` method that prints a summary of the object to the console and a `plot()` method that creates a plot of the object. See `r gh_help("PM_data")` for more information on the `PM_data` class and its methods.

**Note:** We recognize that many users are familiar with the "S3 framework" in R, which uses functions like `summary(object)` and `plot(object)`. To comply with better programming standards, `r pmetrics()` uses the R6 framework. However, we have provided S3 methods for most functions, so you can use `summary(object)` and `plot(object)` if you prefer.

```{r}
#| label: example-npag-6
# S3 method to summarize data
summary(exData)
```

`PM_data` has a `plot()` method that creates a plot of the data. See `r gh_help("plot.PM_data")`  for more information.

```{r}
#| label: example-npag-7
#| eval: false
exData$plot() 
```

```{r}
#| eval: true
#| echo: false
p <- exData$plot()
plotly::plotly_build(p) #needed to display in output
```

### Model objects

You can specify a model by reading a file or directly as a list object in R. We'll do both.

The following code creates the same model as in `/src/model.txt` file. We made the same model in the chapter on [models](models.qmd). The advantage of creating them in R is that one does not need to copy model files into folders to provide necessary inputs.

```{r}
#| label: example-mod1_def
#| eval: true
mod1 <- PM_model$new(
  pri = list(
    Ka = ab(0.1, 0.9),
    Ke = ab(0.001, 0.1),
    V = ab(30, 120),
    lag1 = ab(0, 4)
  ),
  cov = list(
    wt = interp(),
    africa = interp("none"),
    age = interp(),
    gender = interp("none"),
    height = interp()
  ),
  eqn = function(){
    two_comp_bolus
  },
  lag = function(){
    lag[1] = lag1
  },
  out = function(){
  Y[1] = X[2]/V
  },
  err = list(
    proportional(5, c(0.02, 0.05, -0.0002, 0))
  )
)
```




### Fit the model to the data

To keep everything tidy, we are working in a folder specifically to store the runs. We are specifying the path in the call to `$fit()` rather than globally changing the working directory.

```{r}
#| eval: false
#| label: example-npag-8
run1 <- mod1$fit(data = exData, run = 1, overwrite = TRUE, path = "Data/Runs") 
```

The `r gh_help("PM_model")` class has a `$fit()` method that fits the model to the data.
```{r}
#| eval: true
#| echo: false
#| label: example-npag-8-1
run1 <- NPex
```

After the run is complete the results are returned to the assigned object, here 'run1'. However, if you want to load the results later, you can use the `PM_load()` function. Runs will be sequentially numbered as /1, /2, /3,... in your working directory.

```{r}
#| eval: false
#| label: example-npag-9
run1 <- PM_load(1, path = "Data/Runs") # load the results from run 1 if returning to this script later
```

