---
title: "Fit a model with NPAG"
format:
  html:
    include-in-header: Extras/copy-all-code.html
---


```{r}
#| label: setup
#| echo: false
#| message: false
#| eval: true
library(glue)
library(knitr)
library(Pmetrics)


knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = TRUE,
  eval = FALSE
)


pmetrics <- function(){
    knitr::asis_output("[Pmetrics]{style=\"color: #841010; font-family: 'Arial', Arial, sans-serif; font-weight: 900;\"}")
}


r_help <- function(pkg, name, fn = NULL) {
    if(is.null(fn)) {fn <- name}
    glue::glue("[`{name}`](https://rdrr.io/pkg/{pkg}/man/{fn}.html)")
}

gh_help <- function(name) {
    glue::glue("[`{name}`](https://lapkb.github.io/Pmetrics/reference/{name}.html)")
}
```

## Set up the tutorial environment

:::{.callout-tip}
Make sure you have run the [tutorial setup code](index.qmd#tutorial-setup) in your R session before copying, pasting and running example code here.
:::


## Load the Pmetrics library

First, remember you must always load the `r pmetrics()` library to access its functions in your R script or Quarto document.

```{r}
#| eval: false
#| echo: true
#| label: example-npag-1
library(Pmetrics)
```


## Create a run

To review, recall that the typical workflow in `r pmetrics()` is as follows:

1.  Create a `r gh_help("PM_data")` object.
2.  Create a `r gh_help("PM_model")` object.
3.  Fit the model to the data with the `$fit()` method of the model object. This returns a `r gh_help("PM_result")` object, which is also saved to the hard drive and can be loaded later with the `r gh_help("PM_load")` function.
4.  Examine the results with `r gh_help("PM_result")` methods and functions and revise as needed to improve the model fit. 

### Data objects

Since `r pmetrics()` always needs data and a model to run, let's create a new data object.

```{r}
#| echo: false
#| eval: true
#| label: make-data
exData <- dataEx
```

```{r}
#| label: example-npag-4
exData <- PM_data$new(data = "src/ex.csv")
```


### Model objects

You can specify a model as a list object in R. We'll do both.

The following code creates the same model as in `/src/model.txt` file. We made the same model in the chapter on [models](models.qmd). The advantage of creating them in R is that one does not need to copy model files into folders to provide necessary inputs.

```{r}
#| label: example-mod1_def
#| eval: true
mod1 <- PM_model$new(
  pri = list(
    Ka = ab(0.1, 0.9),
    Ke = ab(0.001, 0.1),
    V = ab(30, 120),
    lag1 = ab(0, 4)
  ),
  cov = list(
    wt = interp(),
    africa = interp("none"),
    age = interp(),
    gender = interp("none"),
    height = interp()
  ),
  eqn = function(){
    two_comp_bolus
  },
  lag = function(){
    lag[1] = lag1
  },
  out = function(){
  Y[1] = X[2]/V
  },
  err = list(
    proportional(5, c(0.02, 0.05, -0.0002, 0))
  )
)
```


### Fit the model to the data

To keep everything tidy, we are working in a folder specifically to store the runs. We are specifying the path in the call to `$fit()` rather than globally changing the working directory.

```{r}
#| label: example-npag-8
# results will be saved in the Runs folder in your working directory
# we use run = 1 and overwrite to force the results to be saved in /Runs/1
# omitting overwrite = TRUE would trigger a warning if /1 already exists
# without run = 1, the next available run number will be used

run1 <- mod1$fit(data = exData, run = 1, overwrite = TRUE, path = "Runs") 
```

The `r gh_help("PM_model")` class has a `$fit()` method that fits the model to the data.
```{r}
#| eval: true
#| echo: false
#| label: example-npag-8-1
run1 <- NPex
```

After the run is complete the results are returned to the assigned object, here 'run1'. However, if you want to load the results later, you can use the `PM_load()` function. Runs will be sequentially numbered as /1, /2, /3,... in your working directory.

```{r}
#| eval: false
#| label: example-npag-9
run1 <- PM_load(1, path = "Runs") # load the results from run 1 if returning to this script later
```

