---
title: "More complex models"
format:
  html:
    include-in-header: Extras/copy-all-code.html
---


```{r}
#| label: setup
#| echo: false
#| message: false
#| eval: true
library(glue)
library(knitr)
library(Pmetrics)


knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = TRUE,
  eval = FALSE
)


pmetrics <- function(){
    knitr::asis_output("[Pmetrics]{style=\"color: #841010; font-family: 'Arial', Arial, sans-serif; font-weight: 900;\"}")
}


r_help <- function(pkg, name, fn = NULL) {
    if(is.null(fn)) {fn <- name}
    glue::glue("[`{name}`](https://rdrr.io/pkg/{pkg}/man/{fn}.html)")
}

gh_help <- function(name) {
    glue::glue("[`{name}`](https://lapkb.github.io/Pmetrics/reference/{name}.html)")
}

```

## Set up the tutorial environment

:::{.callout-tip}
Make sure you have run the [tutorial setup code](index.qmd#tutorial-setup) in your R session before copying, pasting and running example code here.
:::

## Introduction

Here we provide examples of adding covariates and compartments to models.

## Model with covariates

Let's create another run object, this time using a model that include covariates.

First, recall the model from the [NPAG example](NPAG.qmd).

```{r}
#| label: example-mod1_def
# model with covariates declared but not used
mod1 <- PM_model$new(
  pri = list(
    Ka = ab(0.1, 0.9),
    Ke = ab(0.001, 0.1),
    V = ab(30, 120),
    lag1 = ab(0, 4)
  ),
  cov = list(
    wt = interp(),
    africa = interp("none"),
    age = interp(),
    gender = interp("none"),
    height = interp()
  ),
  eqn = function(){
    two_comp_bolus
  },
  lag = function(){
    lag[1] = lag1
  },
  out = function(){
  Y[1] = X[2]/V
  },
  err = list(
    proportional(5, c(0.02, 0.05, -0.0002, 0))
  )
)
```

Next, we update it to change `V` to an new variable called `V0` in the `PRI` list element. In the `OUT` list element, we include an equation that relates `V0` to normalized weight. **Note** that we could also have included the `V = V0 * (wt/70)` in a `SEC` block to globally define this relationship for any other model list block. In this case, we only use `V` in the `OUT` block, so we define it there locally.

```{r}
#| label: example-mod2_def
# declare covariates AND use them in the model
mod2 <- PM_model$new(
  pri = list(
    Ka = ab(0.1, 0.9),
    Ke = ab(0.001, 0.1),
    V0 = ab(30, 120),
    lag1 = ab(0, 4)
  ),
  cov = list(
    wt = interp(),
    africa = interp("none"),
    age = interp(),
    gender = interp("none"),
    height = interp()
  ),
  eqn = function(){
    two_comp_bolus
  },
  lag = function(){
    lag[1] = lag1
  },
  out = function(){
    V = V0 * (wt/70)
    Y[1] = X[2]/V
  },
  err = list(
    proportional(5, c(0.02, 0.05, -0.0002, 0))
  )
)
```

Ensure the data are loaded.

```{r}
exData <- PM_data$new(data = "src/ex.csv")
```

Now, run that new model.

```{r}
run2 <- mod2$fit(data = exData, run = 2, path = "Runs", overwrite = TRUE)
```

Right after the run, the results are in `run2`, but for future loading if you return to your script later, you can load it with `PM_load()`.

```{r}
run2 <- PM_load(2, path = "Runs")
```

You can examine the results as we've seen in previous chapters.

## Add compartments to a model

Now, let's create a model with an additional compartment. We'll add a peripheral compartment to the previous two-compartment bolus model.

```{r}
#| label: example-mod3_def
# model with covariates and an additional compartment
mod3 <- PM_model$new(
  pri = list(
    Ka = ab(0.1, 0.9),
    Ke = ab(0.001, 0.1),
    V0 = ab(30, 120),
    KCP = ab(0.001, 0.5),
    KPC = ab(0.001, 0.5),
    lag1 = ab(0, 4)
  ),
  cov = list(
    wt = interp(),
    africa = interp("none"),
    age = interp(),
    gender = interp("none"),
    height = interp()
  ),
  eqn = function(){
    three_comp_bolus
  },
  lag = function(){
    lag[1] = lag1
  },
  out = function(){
    V = V0 * (wt/70)
    Y[1] = X[2]/V
  },
  err = list(
    proportional(5, c(0.02, 0.05, -0.0002, 0))
  )
)
```

**Notes:**

1. We used `r gh_help("model_lib")` to check the name of the algebraic 3-compartment model and the required parameters that should be declared (Ka, Ke, KCP, KPC and V).
2. Don't forget that you can simply type the name of the library model into the console to get information about it.
```{r}
# type the name to get information about the model
three_comp_bolus
```
3. Output is still in compartment 2.



Now, run that new model.
```{r}
# run model with 3 compartments and covariates
run3 <- mod3$fit(data = exData, run = 3, path = "Runs", overwrite = TRUE)
```

Again, you can load it later with `PM_load()`.

```{r}
run3 <- PM_load(3, path = "Runs")
```

## Differential equations

Here is `mod3` again, but this time using differential equations instead of built-in compartmental models, just to show the structure of the ODEs.

```{r}
#| label: example-mod4_def
# ODE model with covariates and an additional compartment
mod3b <- PM_model$new(
  pri = list(
    Ka = ab(0.1, 0.9),
    Ke = ab(0.001, 0.1),
    V0 = ab(30, 120),
    K12 = ab(0.001, 0.5),
    K21 = ab(0.001, 0.5),
    lag1 = ab(0, 4)
  ),
  cov = list(
    wt = interp(),
    africa = interp("none"),
    age = interp(),
    gender = interp("none"),
    height = interp()
  ),
  eqn = function(){
    dX[1] = B[1] - Ka * X[1]
    dX[2] = R[1] + Ka * X[1] - (K12 + Ke) * X[2] + K21 * X[3]
    dX[3] = K12 * X[2] - K21 * X[3]
  },
  lag = function(){
    lag[1] = lag1
  },
  out = function(){
    V = V0 * (wt/70)
    Y[1] = X[2]/V
  },
  err = list(
    proportional(5, c(0.02, 0.05, -0.0002, 0))
  )
)
```

**Notes:**

1. We have replaced `KCP` and `KPC` with `K12` and `K21`, to emphasize that ODE models do not require specifically named parameters in the `PRI` block. 
2. By including `R[1]` in the central compartment, `X[2]` in this model, we allow for infusions of drug 1 in addition to the bolus input for drug 1, `B[1]`. The equivalent algebraic model, `three_comp_bolus`, does the same.