
# Probability of Target Attainment (PTA) Analysis

::: {.callout-tip} 
## Learn more
It is essential that you understand simulations first. If necessary, refer to [Simulation](Simulate.qmd). 
:::



**Note**: these can be computationally intense and take some time.

Be sure to have executed the NPAG run above and used PM_load(2) in EXERCISE 2
Type ?PM_sim, ?PM_pta, or ?plot.PM_pta into the R console for help.

simulate with the template data file that contains different doses
Look at PM_sim for help on arguments to this function, including predInt,
seed, limits, nsim.


```{r}
#| eval: false
#| echo: true

simlist1 <- run1$sim(
  limits = c(0, 3), data = "../src/ptaex1.csv",
  predInt = c(120, 144, 0.5), seed = rep(-17, 4)
)
```

now simulate with covariates; make sure that you defined the covariate
object first in Exercise 5 above and have loaded the results of Exercise 2
with PM_load(2)

```{r}
#| eval: false
#| echo: true
#| 
simlist2 <- run2$sim(
  limits = 5, data = "../src/ptaex1.csv",
  predInt = c(120, 144, 0.5), seed = rep(-17, 4),
  covariate = covariate
)
```

make the first PMpta object to calculate the time above each target for at
least 60% of the dosing
interval from 120 to 144 hours.  Include labels for the simulations.
?makePTA for help
define simulation labels first

```{r}
#| eval: false
#| echo: true

simlabels <- c("600 mg daily", "1200 mg daily", "300 mg bid", "600 mg bid")

pta1_2 <- PM_pta$new(
  simdata = simlist1,
  target = c(0.25, 0.5, 1, 2, 4, 8, 16, 32), target_type = "time",
  success = 0.6, start = 120, end = 144
)

pta1b_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = simlabels,
  target = c(0.25, 0.5, 1, 2, 4, 8, 16, 32), target_type = "time",
  success = 0.6, start = 120, end = 144
)

# summarize the results
pta1_2$summary()
```


in the summary()$pta, reg_num is the simulation template ID number;
target in this case is the MIC; prop_success is the proportion of the simulated
profiles for each dose/MIC that are above the success threshold (0.6); pdi.mean and pdi.sd
are the mean and standard deviation of the pharmacodynamic index (PDI), in this case proportion of the interval > MIC.
In the $pdi, target and simnum are the same, but now the median and confidence
intervals (default 95%) PDI are shown.
?summary.PMpta for help

Plot the first without covariates.   We didn't include simulation
labels in the makePTA command, so generics are used here, but we move it to
the bottom left; ?legend for help on arguments to supply to the
legend list argument to plot.PMpta.

```{r}
#| eval: false
#| echo: true

pta1_2$plot(ylab = "Proportion with %T>MIC of at least 60%", grid = TRUE, legend = list(x = "bottomleft"))

pta1b_2$summary()
```

Plot the second with covariates simulated. Note the regimen labels are included, but we move
the legend to the bottom left.

```{r}
#| eval: false
#| echo: true

pta1b_2$plot(
  ylab = "Proportion with %T>MIC of at least 60%", grid = TRUE,
  legend = list(x = "bottomleft")
)

# Now we'll define success as free auc:mic > 100 with a free drug fraction of 50%
pta2_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = simlabels, target = c(0.25, 0.5, 1, 2, 4, 8, 16, 32),
  free_fraction = 0.7,
  target_type = "auc", success = 100, start = 120, end = 144
)

summary(pta2_2)
pta2_2$plot(
  ylab = "Proportion with AUC/MIC of at least 100", grid = TRUE,
  legend = list(x = "bottomleft")
)

# success is Cmax/MIC >=10
pta3_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = simlabels,
  target = c(0.25, 0.5, 1, 2, 4, 8, 16, 32),
  target_type = "peak", success = 10, start = 120, end = 144
)

pta3_2$summary()
pta3_2$plot(ylab = "Proportion with peak/MIC of at least 10", grid = TRUE)

# success = Cmin:MIC > 1
pta4_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = simlabels,
  target = c(0.25, 0.5, 1, 2, 4, 8, 16, 32),
  target_type = "min", success = 1, start = 120, end = 144
)

pta4_2$summary()
pta4_2$plot(ylab = "Proportion with Cmin/MIC of at least 1", grid = TRUE, legend = list(x = "bottomleft"))

# now plot the PDI (pharmacodynamic index) of each regimen, rather than the proportion
# of successful profiles.  A PDI plot is always available for PMpta objects.
pta4_2$plot(at = 1, type = "pdi", ylab = "Cmin:MIC", grid = TRUE)

# Each regimen has the 90% confidence interval PDI around the median curve,
# in the corresponding, semi-transparent color.  Make the CI much narrower...
pta4_2$plot(at = 1, type = "pdi", ci = 0.1)

# ...or gone altogether, put back the grid, redefine the colors, and make lines narrower
pta4_2$plot(
  at = 1, type = "pdi", ci = 0, grid = TRUE,
  line = list(
    color = c("blue", "purple", "black", "brown"),
    width = 1
  )
)
```


now let's repeat the analysis but simulate the distribution of MICs
using susceptibility of Staphylococcus aureus to vancomycin contained
in the mic1 dataset within Pmetrics (?mic1)

```{r}
#| eval: false
#| echo: true

# see the source with ?mic1
pta4b_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = c("600 mg daily", "1200 mg daily", "300 mg bid", "600 mg bid"),
  target = makePTAtarget(mic1), target_type = "min", success = 1, start = 120, end = 144
)

pta4b_2$summary()
# plot it
pta4b_2$plot(
  grid = TRUE, ylab = "Proportion with Cmin/MIC of at least 1",
  marker = list(color = "red"), line = list(color = "black")
)
pta4b_2$plot(type = "pdi", grid = TRUE, ylab = "Proportion with Cmin/MIC of at least 1")
```

note that the plot changes since target MICs are no longer discrete
since most of the MICs are very low, the regimens all look very similar

```{r}
#| eval: false
#| echo: true

# success = concentration at time 3 hours:MIC > 2
pta5_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = simlabels,
  target = c(0.25, 0.5, 1, 2, 4, 8, 16, 32), target_type = 123, success = 2, start = 120, end = 144
)
pta5_2$summary()
pta5_2$plot(ylab = "Proportion with C3/MIC of at least 1", grid = TRUE, legend = list(x = .3, y = 0.1))


# success is trough >10
pta6_2 <- PM_pta$new(
  simdata = simlist2,
  simlabels = simlabels,
  target = 10, target_type = 144, success = 1, start = 120, end = 144
)
plot(pta6_2)
pta6_2$summary()
```
